\speculake: Hiding Malware in Speculative Execution

### 2018-05-09 paper/discussion.tex
discussion pass

\subsection{Defenses}
We now address possible defenses to this new class of malware introduced by
\speculake. Though we caution that these defenses often have drawbacks. The core
component of \speculake is taking advantage of speculative execution inherent in
CPUs today, thus these defenses focus on limiting this behavior.

\subsubsection{Operating System Defenses}
One defense to limit the capabilities of \speculake is to introduce privileged
instructions that grant the Operating System more control.

A necessary component of \speculake is the ability of a process to affect the
Branch History Buffer (BHB), which is used by all processes, allowing programs
to affect how separate programs choose to make indirect jumps.

To address this a new set of instructions allowing the OS to set and clear the
Branch History Buffer on context switches allows for processes to maintain a
useful branch history while disabling other processes from affecting their jump
decisions.

By design these new instructions grant the operating system control over the
state a program generates introducing possible attack vectors for malicious
Operating Systems to poison the Branch History Buffer or even allow \speculake
type attacks.

\subsubsection{Malware Detection}
\label{subsubsec:malware}
The world of computer security is often referred to as an arms race, hackers and
criminals racing to create and use exploits before security researchers and
software engineers can create defenses. As such we expect to see the same type
of race with this new class of malware.

A necessary portion of this class of malware is utilizing a cache side-channel,
determining how long certain values require to load from memory or cache. Thus
one method of defense against \speculake is to adapt malware detectors to
examine if a program checks for the timing of loading values from memory.

Initially this appears to be a very effective method of recognizing this attack.
However, the notion of timing memory loads can be done in obscure ways which can
be made to appear innocuous. As opposed to directly measuring the timing of
memory loads, the timing can be more indirect: measuring how long a specific
loop takes that happens to do a memory load. Or more directly, avoiding
measuring time altogether: create a race-condition which only the
already-in-cache value will easily win.

While the notion of simply adapting current defenses to recognize this new
strain of malware may be simplest, we caution that the methods introduced by
\speculake are very general and can be accomplished in numerous ways and simply
attempting to recognize one method will undoubtedly be circumvented by future
generations of this malware.

\subsubsection{Cache Poisoning}
As mentioned in Section~\ref{subsubsec:malware} a vital component of \speculake
is the ability to perform cache timings. Any mechanism that modifies the cache
can affect the success of \speculake-like attacks.

Drastic methods such as completely poisoning the cache will hinder this attack
while also forcing almost all loads to result in a cache miss, effectively
nullifying the point of cache.

However, more conservative methods such as randomly loading values into cache
can have a negative affect on \speculake-like attacks (causing the cache timing
to predict incorrect values) while not having too detrimental of an affect on
programs as a whole.

In response to this method of defense \speculake-like attacks can modify the
number of iterations and confidence thresholds necessary to determine a value.
As mentioned in Section~\ref{subsec:impl-turing} our experiments found that it
was only necessary to perform 50 iterations of our \texttt{target\_fn} in order
to gain 80\% confidence in the value returned. While only 50 iterations are
necessary, increasing this value to higher levels can lead to a significantly
higher confidence with very little performance detriment (especially in the
case of AES decryption).

\subsection{Resilience of \speculake}
Spectre works to use a cache side-channel to leak information from a process. As
such defenses to Spectre work to limit the information available to a
speculatively executing program. One such solution may be a new instruction such
as  \texttt{specfence}: an instruction which designates to the CPU a point that
cannot be speculated past. This would require that loads/stores after the
\texttt{specfence} instruction would be on hold until loads/stores from before
the \texttt{specfence} instruction have retired. The loads/stores after a
\texttt{specfence} do not begin to execute at all until the loads/stores before
the \texttt{specfence} retire.

However effective such an instruction may be against Spectre, it is not a
defense against \speculake. Such an instruction would be useful in restricting
the execution of a program to restrict the speculative capabilities of program,
thereby stopping information from being leaked. However, a malicious user
creating \speculake malware would simply choose to not use this instruction.

This demonstrates the versatility of speculative execution in general. While
defenses against Spectre are needed, a more general purpose solution to
speculative execution as a whole is necessary to address the wide range of its
capabilities.

% We note that this instruction seems to already exist, albeit unintentionally.
% The \texttt{rep} instruction repeats the execution of string instructions a set
% number of times (set in a register). When set to repeat 90 or more times, these
% instructions are not speculated past, requiring they complete and be retired
% before looking to the next instruction. This appears to be due to the 
% \texttt{rep} instruction adding additional $\mu$ ops per instruction, we found
% that using the \texttt{rep} instruction caused $2n$ $\mu$ ops when the string
% operation works on $n$ bytes~\cite{uop_article}. Thus enough repetitions would cause instructions
% after the \texttt{rep} instruction to wait for the full repetition to be retired
% before beginning their execution.

@@

\subsubsection{Multiple Triggers}
To create further difficulties for an analyst, or to further target the
execution environment it is possible to have the payload program test a series
of separate jump patterns. The decryption process can then be done in stages
where the result of a given stage determines the initial state of the next stage
that the payload program performs. Only when all triggers have been completed in
the correct order will the true plaintext be available.

While this would increase the difficulty for analyst  who has access to a sample
in isolation, it  does not add any extra layers of complication for an analyst
who can observe the entire decryption process. Any state encoded in the real
world during the individual stages will be available to current debugging tools.
This allows the initial state of a sample to be modified to simplify subsequent
runs. For example, they could patch the binary to begin with the state of the
final stage such that only the final trigger must be run again to retreive the
plaintext. While this simplifies the job of acquiring the plaintext of a given
\speculake sample, it does not give any information about the trigger programs
or the key schedules used.

@@

Alternatively multiple triggers could be used make individual state changes in
the payload program, and write in data from a local, or (as demonstrated in
section~\ref{subsec:openssl}) remote processes. For example, a payload program
can check a series of three distinct jump patterns corresponding to three
different triggers. If the first trigger is detected the payload shifts in a 1
to the state; if the second is detected it shifts in a 0. If the third trigger
is detected it discontinues listening to the three triggers and proceeds with
other computation. This could be a direct form of \speculake using emulation and
decryption or more traditional malicious behavior. Using this method a payload
program can be distributed without the encrypted data section all together,
instead having the data written into the process at run-time without ever
connecting out to any other resource.

\subsubsection{Benign Triggers}
We have demonstrated that a using OpenSSL as a benign trigger allows for
\speculake payloads to be triggered remotely, and discussed the minimum
requirements for identifying a benign trigger. However, we have not classified
typical system resources that might be used a benign triggers. This is a
promising direction for future investigation as various system libraries present
juicy targets for trigger design. For exmaple, if a sufficient series of jumps
is found to exist in the TCP stack implemented in the unix kernel (i.e. in the
TCP Checksum library) an attacker could reasonably trigger a \speculake payload
remotely, from a spoofed IP address. This attack would rely on a payload
defeating KASLR, however, this has been demonstrated to be
feasible~\cite{gruss2017kaslr,evtyushkin2016jump}

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this
model of hidden computation applies more broadly. As of now there are no ways of
debugging speculative computation meaning that the computation done there is
effectively isolated as with trusted execution environments. The security
provided by the obfuscated key schedule makes \speculake useful in the domains
of Digital Rights Management (DRM) and whitebox cryptography, and approximate
computing fields. 
%% Approximate Computing

Approximate Computing prioritizes the computation of quick results that are
potentially incorrect. This form of computation often leverages some type of
machine learning to determine how well individual results fit with a crowd of
approximately computed results. As mentioned in Section~\ref{subsec:impl-turing}
occasionally values necessary for computation in the speculative world are not
in cache and thus must be speculated themselves. In this situation error
correcting codes are useless given that the mistake is not a mutation of the
correct input, but rather a guess at what the input might be. The mechanisms
used in approximate computing may be beneficial here, allowing computation to
run speculatively and results be analyzed after the computation to determine how
results fit, with the understanding that these results may not be accurate.

>>>>

\subsection{Defenses}
We now address possible defenses to detecting and reverse engineering malware that uses
\speculake.

\subsubsection{Malware Detection}
\label{subsubsec:malware}
Like Spectre, \speculake relies on a cache side-channel to transfer
information from the speculative world to the real world. Since the payload
program much at least occasionally watch this side channel, it offers a
potential method for detecting \speculake malware.

Analysts could search for telltale signs of cache inference behavior, such as
the use of \texttt{clflush} instructions or reading cycle timings. However,
while this would allow detection of \speculake malware, it does not directly
help in reverse engineering the malware's behavior. We also note that our
implementation of \speculake used the \texttt{Flush+Reload} cache side channel
method, but other methods (such as \texttt{Prime+Probe}) could be used in its
place.

In addition, the malware can also employ more subtle and less direct methods
of measuring what is in the cache. For example, creating race conditions between
multiple threads that results in different behaviors depending on which threads
values are cached.

Finally, \speculake could also use another side channel method besides that
avoids the cache
to exfiltrate information from the speculative gadgets, such as the branch
predictor itself~\cite{evtyushkin2018branchscope}, memory bandwidth, or power
utilization of the system.

%While the notion of simply adapting current defenses to recognize this new
%strain of malware may be simplest, we caution that the methods introduced by
%\speculake are very general and can be accomplished in numerous ways and simply
%attempting to recognize one method will undoubtedly be circumvented by future
%generations of this malware.

\subsubsection{Hardware-level defenses}
Ultimately, speculative execution vulnerabilities stem from problems in
hardware, and are best fixed there. However, we note that fixing problems in
order to prevent Spectre may not necessarily prevent \speculake malware. In Spectre, an application is
coerced against its will to mis-speculate branches, but in \speculake, the
application willfully encourages the mis-speculation. Thus, some solutions may
defend against Spectre, but not \speculake.

One such solution is a new \texttt{specfence} instruction, that prohibits the
CPU from speculating past. This instruction would require loads and stores after
the instruction must wait (and cannot be pre-fetched) until the
\texttt{specfence} instruction has retired. We note this is distinct from the
\texttt{mfence} instruction that serializes memory transactions. Speculative
reads and out-of-order execution can still occur past an \texttt{mfence}, and
the instruction does not serialize the instruction
stream~\cite{intel-software-guide}.

One way to achieve a \texttt{specfence} pseudo-instruction is to fill the ROB
with enough instructions that the CPU cannot speculate past it. We find that the
\texttt{rep} instruction is able to do this if it is set to repeat more than 96
times, presumably as it issues $2*96$ $\mu$-ops, and fills the
ROB~\cite{intel-instruction-tables}.

However effective such an instruction may be against Spectre, it is not a
defense against \speculake. A malicious \speculake malware author would simply
not include such an instruction in either their payload or trigger programs.
% Could add:
%This demonstrates the versatility of speculative execution in general. While
%defenses against Spectre are needed, a more general purpose solution to
%speculative execution as a whole is necessary to address the wide range of its
%capabilities.

\paragraph{Operating System Defense}
Hardware could also provide the operating system with a privileged instruction
that gives it more control over the interaction between programs. For example,
the OS could use a privileged instruction to copy out the Branch History Buffer
(BHB) state when it context switches between processes, restoring it when the
process is restored. This would prevent cross-process interactions with the
branch predictor, and prevent multi-process \speculake malware. However, it does
not prevent a single-process \speculake, and it does not fully address the
Spectre vulnerability.

%\subsubsection{Cache Poisoning}
%As mentioned in Section~\ref{subsubsec:malware} a vital component of \speculake
%is the ability to perform cache timings. Any mechanism that modifies the cache
%can affect the success of \speculake-like attacks.
%
%Drastic methods such as completely poisoning the cache will hinder this attack
%while also forcing almost all loads to result in a cache miss, effectively
%nullifying the point of cache.
%
%However, more conservative methods such as randomly loading values into cache
%can have a negative affect on \speculake-like attacks (causing the cache timing
%to predict incorrect values) while not having too detrimental of an affect on
%programs as a whole.
%
%In response to this method of defense \speculake-like attacks can modify the
%number of iterations and confidence thresholds necessary to determine a value.
%As mentioned in Section~\ref{subsec:impl-turing} our experiments found that it
%was only necessary to perform 50 iterations of our \texttt{target\_fn} in order
%to gain 80\% confidence in the value returned. While only 50 iterations are
%necessary, increasing this value to higher levels can lead to a significantly
%higher confidence with very little performance detriment (especially in the
%case of AES decryption).

\subsection{Resilience of \speculake}
%Spectre works to use a cache side-channel to leak information from a process. As
%such defenses to Spectre work to limit the information available to a
%speculatively executing program. 
%

\subsubsection{Gadget Location}
The benefits of \speculake rely on the computation performed in the
speculative function being well hidden. To this end we have verified 
certain requirements about the locations that a speculative gadget can 
be placed within a binary.

First and foremost the gadget must be in the instruction cache. 
This does not require that the actual instructions ever run,
 only that they are in cache. We have 
found that this is not a large barrier to overcome as Intel 
processors aggressively prefetch cache pages~\cite{measuring-cache}
and simple access patterns could be established to ensure
that the required block(s) are accessible.

We have also confirmed that all speculative code in the payload must be
stored in executable pages, and that it cannot be stored in non-executable data
sections. This is somewhat surprising, given the nature of the Meltdown
vulnerability that enabled speculative execution past page faults. However, in
the case of Meltdown, the speculative computation still occurs in pages marked
as executable.

%We suspect that this is again due to caching. The processor has no
%reason to cache instructions stored in the data section in the instruction
%cache. If a trigger trains the branch predictor to jump to an address in the
%payload's data section, the payload will not be able to fetch the instructions
%into the execution pipeline.

\subsubsection{Hypervisors}
For virtual environments that provide guests significant periods of access to 
hardware resources, we have found that \speculake is viable in certain configurations.

\texttt{Guest1 -> Guest1} - When both trigger and payload are run internal to
a single virtual machine the branch prediction is successfully trained in the payload
program, though at a notably reduced hit rate. This can be overcome by using an 
increased number of iterations to establish confidence.

\texttt{Guest -> Host} - In a configuration where trigger runs on the guest while the 
payload runs on the host we have found that \speculake also successfully trains the 
branch predictor well enough retreive a signal. Again, this signal is significantly
reduced, implying the hypercall context swtich from guest to host requires some 
overhead. We note that processor affinity in a guest is not mapped to processor 
affinity in the host, as the host assigns the vm to any available CPU, though 
eventually the hypervisor will assign the vm back to the same core as the payload 
program allowing progress to continue.

\texttt{Host -> Guest} - When trigger runs on the host attempting to train the 
branch predictor to influence a payload process in a guest we have found that 
the payload receives no signal. This implies that the hypercall context switch 
from the host to the guest is heavier and has a larger impact on branch history. 
This matches with work done by the Project Zero team at Google which found that 
hypervisors can be fingerprinted from within a vm using the branch 
predictor~\cite{project_zero}.

\texttt{Guest1 -> Guest2} - Configurations where trigger and payload 
run on coresident virtual machines similarly show no signal for the same 
reasons that host to guest \speculake fails. However, this has the added difficulty
that virtual machines must be run on the same core or on a pair of 
hyperthreaded cores.

@@

\subsubsection{Multiple Triggers}
To create further difficulties for an analyst, or to further target the
execution environment it is possible to have the payload program test a series
of separate jump patterns. Instead of requiring only a single trigger program,
the payload could require multiple trigger programs to be running
simultaneously, or in a particular order.

This could allow fine-grained targeting of malware. For instance, the attacker
could distribute trigger programs to different channels to target different sets
of victims, and have the ultimate payload only operate at the intersection of
these groups. As an example, one trigger program could be distributed to a
particular country (e.g. Iran), and another to a particular device globally
(centrifuge controllers), resulting in the malicious payload (Stuxnet) only
being revealed and executed on the intersection of these two groups.

@@

Alternatively multiple triggers could be used make individual state changes in
the payload program, and write in data from a local, or (as demonstrated in
section~\ref{subsec:openssl}) remote processes. For example, a payload program
can check a series of three distinct jump patterns corresponding to three
different triggers. If the first trigger is detected the payload shifts in a 1
to the state; if the second is detected it shifts in a 0. If the third trigger
is detected it discontinues listening to the three triggers and executes the
data shifted into the state.
This could be a direct form of \speculake using emulation and
decryption or more traditional malicious behavior. Using this method a payload
program can be distributed without the encrypted data section all together,
instead having the data written into the process at run-time without ever
connecting out to any other resource.

\subsubsection{Benign Triggers}
We have demonstrated that a using OpenSSL as a benign trigger allows for
\speculake payloads to be triggered remotely, and discussed the minimum
requirements for identifying a benign trigger. However, we have not classified
typical system resources that might be used as benign triggers. This is a
promising direction for future investigation as various system libraries present
juicy targets for trigger design. For exmaple, if a sufficient series of jumps
is found to exist in the TCP stack implemented in the unix kernel (i.e. in the
TCP Checksum library) an attacker could reasonably trigger a \speculake payload
remotely, from a spoofed IP address. This attack would rely on a payload
defeating KASLR, however, this has been demonstrated to be
feasible~\cite{gruss2017kaslr,evtyushkin2016jump}

%% Approximate Computing 
\subsubsection{Approximate Computing}
Approximate Computing prioritizes the computation of quick results that are
potentially incorrect. This form of computation often leverages some type of
machine learning to determine how well individual results fit with a crowd of
approximately computed results. As mentioned in Section~\ref{subsec:impl-turing}
occasionally values necessary for computation in the speculative world are not
in cache and thus must be speculated themselves. In this situation error
correcting codes are useless given that the mistake is not a mutation of the
correct input, but rather a guess at what the input might be. The mechanisms
used in approximate computing may be beneficial here, allowing computation to
run speculatively and have an outer process that is tolerant of errors or
mis-steps by the speculative world.

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this
model of hidden computation applies more broadly. As of now there are no ways of
debugging speculative computation, meaning that the computation done there is
effectively isolated and difficult to extract by third-parties. While not a true
trusted execution environment, it could be used like one, offering potential
uses in Digitial Rights Management (DRM), and whitebox cryptography.
